---
title: "Modelling Data Streams with R"
author: 'Author: Dickson Owuor (Ph.D.)'
date: "13/06/2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modelling Data Streams with R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## About Data streams

A data stream is a continuous sequence of data points generated from a non-stationary process. In this tutorial, we learn how to use **stream** R library to simulate and model data streams. The tutorial is adopted from these works:

-   [R package stream](https://github.com/mhahsler/stream)

-   [MLPack for Hoeffding Tree](https://search.r-project.org/CRAN/refmans/mlpack/html/hoeffding_tree.html)

## Outline

1.  Installation of libraries
2.  Usage and documentation
3.  Setting up data streams and visualization
4.  Clustering data streams
5.  Writing and reading data streams from files
6.  Classifying data streams (assignment)

### 1. Installing libraries

Note: un-comment the lines if the libraries ARE NOT installed on your system.

```{r eval=FALSE}
#install.packages("stream")
#install.packages("animation") # for animating data streams
#install.packages("mlpack") # for hoeffding-tree algorithm
any(grepl("stream", installed.packages())) # Check if installed (TRUE)

# Load the libraries
library("stream")
library("mlpack")

```

### 2. Usage and documentation

In order to find out about the **stream** R library, we can use the *help* command.

```{r eval=FALSE}
??stream # Use help to check out the documentation
```

Here is the [CRAN URL](https://cran.r-project.org/web/packages/stream/index.html).

### 3. Setting up data streams and visualization

We can create dummy data streams from functions such as *DSD_Benchmark* or from real world data sets (i.e., *cassini*).

```{r eval=FALSE}
stream1 <- DSD_Benchmark(1) 
animate_data(stream1, horizon=100, n=5000, xlim=c(0,1), ylim=c(0,1))

# Cassini dataset
stream2 <- DSD_mlbenchGenerator("cassini")
animate_data(stream2)
```

### 4. Clustering data streams

In order to cluster data streams, we first create a random data stream with 3 Gaussian clusters and 10% noise.

```{r eval=FALSE}
set.seed(2000)
stream3 <- DSD_Gaussians(k = 3, noise = 0.1)

plot(stream3)

```

Next, we cluster a stream of 1000 points using *D-Stream* function.

```{r eval=FALSE}
dstream <- DSC_DStream(gridsize = 0.1)
update(dstream, stream3, 1000)

plot(dstream, stream3, grid = TRUE)

```

### 5. Writing and reading data streams from files

First, we create a data stream and write it to disk.

```{r eval=FALSE}
stream4 <- DSD_Gaussians(k=3, d=5, outliers=1, space_limit=c(0,2), outlier_options = list(outlier_horizon=10))

write_stream(stream4, "data.txt", n=10, header = TRUE, sep=",", class=TRUE, write_outliers=TRUE)


```

Next, we read the same data stream back (as a loop)

```{r eval=FALSE}
stream5 <- DSD_ReadCSV(k=3, o=1, "data.txt", sep=",", header = TRUE, loop=TRUE, class="class", outlier="outlier") 
animate_data(stream2)

```

### 6. Classifying data streams (assignment)

The last exercise in this tutorial involves classifying data streams using the **hoeffding-tree**. We use the *MLPack* R library: [MLPack CRAN Library](https://search.r-project.org/CRAN/refmans/mlpack/html/hoeffding_tree.html).

This exercise is an assignment in which you are required to:

a.  Identify the error in the lines of code below, and

b.  rectify them to get the correct output.

```{r eval=FALSE}
# Documentation
??hoeffding_tree

# Classification
output <- mlpack::hoeffding_tree(training=stream2, confidence=0.99)
tree <- output$output_model

```
