---
title: "R-MariaDB Tutorial"
author: "Author: George Obaido"
date: "14/07/2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing SQL with dbplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### About RMariaDB Package
RMariaDB is a database interface and MariaDB driver for R. This version is aimed at full compliance with the DBI specification, as a replacement for the old RMySQL package. For more information, visit RMariaDB’s official site: <http://rmariadb.r-dbi.org>

### Installation
To install from CRAN:
```{r}
#install.packages("RMariaDB")
#install.packages("DBI")

library(RMariaDB)
library(DBI)
library(dplyr)
library(dbplyr)
```

### Creating a connection
There are 5 settings needed to make a connection:

1. Driver - See the Drivers section for setup information
2. Server - A network path to the database server
3. UID - User name used to access MySQL server
4. PWD - The password corresponding to the provided UID
5. Port - Should be set to 3306

```{r}
con <- dbConnect(
  drv = RMariaDB::MariaDB(), 
  username = "root",
  password = "", 
  host = "localhost", 
  port = 3306
)
```

### R Commands for databases
R uses a number of commands for databases, as follows:

1. The default ***dbCreateTable()*** method calls sqlCreateTable() and dbExecute().
2. ***dbExecute***: Execute an update statement, query number of rows affected, and then close result set.
3. ***dbGetQuery()*** always returns a data.frame with as many rows as records were fetched and as many columns as fields in the result set, even if the result is a single value or has one or zero rows.

A comprehensive list of commands can be found in: [https://rdrr.io/cran/DBI/](https://rdrr.io/cran/DBI/) 


### Getting started
Connect and use a database,this will always return an empty string :

```{r}
dbExecute(con, "USE northwind;")
```

Return all data in the db table.
```{sql, connection=con, max.print = 20}
SELECT * FROM customers
```
### Creating a Database
```{sql, connection=con}
CREATE DATABASE IF NOT EXISTS testDB;
```

### Initiate the database
Connect and use a database,this will always return an empty string:

```{r}
dbExecute(con, "USE testDB;")
```

### Creating a table
The following example creates a table called "Persons" that contains five columns: PersonID, LastName, FirstName, Address, and City:

The datatype parameter specifies the type of data the column can hold (e.g. varchar, integer, date, etc.).
```{sql, connection=con}
CREATE TABLE IF NOT EXISTS Persons(
    PersonID int(11) NOT NULL AUTO_INCREMENT PRIMARY KEY,
    LastName varchar(255),
    FirstName varchar(255),
    Address varchar(255),
    City varchar(255)
)
```


### Inserting into a table
To add multiple rows to a table at once, you use the following form of the INSERT statement:

```{sql, connection=con}
INSERT IGNORE INTO Persons (PersonID, LastName, FirstName, Address, City)
VALUES 
(1,'Erichsen','Tom ','1 Stavanger 4006','Oslo'),
(2,'Philips','James ','21 Folks Rd','Johannesburg'),
(3,'Lee','Kim ','2 Greg Rd','Nairobi'),
(4,'Doe','John ','55 Yawn Rd','Pretoria'),
(5,'Marks','Cindy ','22 Folks Rd','Abuja');
```


### Retrieve records from a table
The following SQL statement fetches all the columns from the "Persons" table:
```{sql, connection=con, tab.cap = "Person's table"}
SELECT * FROM Persons
```
### Update records in a table
The following SQL statement updates the first customer (PersonID = 1) with a new last name and a new city.
```{sql, connection=con}
UPDATE Persons
SET LastName = 'Schmidt', City= 'Frankfurt'
WHERE PersonID = 1;
```


### Using Aliases
The following SQL statement creates two aliases, one for the CustomerName column and one for the ContactName column. Note: It requires double quotation marks if the alias name contains spaces:

```{sql, connection=con}
SELECT LastName AS "Surname" FROM Persons;
```
