---
title: "Data Modelling with SQL using R"
author: 'Author: Dickson Owuor (Ph.D.)'
date: "06/07/2022"
output: 
  html_notebook:
    toc: yes
    number_sections: false
---

# Structured Query Language (SQL)
SQL is a standardized programming language that is used to manage relational databases and perform various operations on the data in them.

**RMariaDB** is a database interface and MariaDB driver for R. This version is aimed at full compliance with the DBI specification, as a replacement for the old RMySQL package. For more information, visit RMariaDB’s official site: <http://rmariadb.r-dbi.org>. Apart from the **RMariaDB** package library, we also use the following libraries:

* [CRAN DBI library](https://cran.r-project.org/web/packages/DBI/index.html): provides a database interface definition for communication between R and relational database management systems (in this case MariaDB).
* [CRAN dbplyr library](): a back end for databases that allows you to work with remote database tables as if they are in-memory data frames.


# Installing MariaDB
Before attempting any task in this tutorial, you need to install **MariaDB Server** in your Computer. Follow the steps that follow.

### Windows and Linux Machines
Direct downloads and installations are available. Follow this link to download and install **MariaDB**:

* <https://www.mariadbtutorial.com/getting-started/install-mariadb/>.

### MacOS Machines
At the time of creation of this tutorial, no options for directly downloading and installing **MariaDB** exist. However, it is possible to install it via **Homebrew** software package management tool.

* Install **Homebrew** via this link <https://brew.sh/>
* After installing Homebrew, **MariaDB Server** can be installed with this command:

```{bash eval=FALSE}
brew install mariadb
```

After installation, start MariaDB Server:

```{bash eval=FALSE}
mariadb  # starts the server, press Ctrl-C to stop it

# Statement to change password of root user
ALTER USER 'root'@'localhost' IDENTIFIED BY 'YOURNEWPASSWORD';

```


# Installation of libraries
Install and load the following package libraries.

```{r eval=FALSE}
install.packages("RMariaDB")
install.packages("DBI")
install.packages("dbplyr")


library(RMariaDB)
library(DBI)
library(dplyr)
library(dbplyr)

```

# Usage and documentation

```{r eval=FALSE}
??RMariaDB

```

# Making a DB connection
In order to connect our R program to a remote database, we need to specify the following 5 configurations:

1. Driver - See the Drivers section for setup information
2. Host (or server) URL - A network path to the database server
3. Username - User name used to access MySQL server
4. Password - The password corresponding to the provided UID
5. Port number - normally it is 3306


```{r eval=FALSE}
?dbConnect
  
con <- dbConnect(
  drv = RMariaDB::MariaDB(), 
  username = "root",
  password = "", 
  host = "localhost", 
  port = 3306
)

```


# R Commands for manipulating DBs
R uses a number of commands for performing numerous tasks on databases (mostly from the **DBI** library). Here are a few commands:

* The default **dbCreateTable** method calls sqlCreateTable() and dbExecute().
* **dbExecute**: Execute an update statement, query number of rows affected, and then close result set.
* **dbGetQuery** always returns a data.frame with as many rows as records were fetched and as many columns as fields in the result set, even if the result is a single value or has one or zero rows.

A comprehensive list of commands can be found in: [https://rdrr.io/cran/DBI/](https://rdrr.io/cran/DBI/)


```{r eval=FALSE}
?dbCreateTable
?dbExecute
?dbGetQuery

```

# Manipulating DBs
The rest of this tutorial guides you on how to invoke various SQL statements from your **R script**.


### Create a DB
Invoke the following statement to create a new database.

```{sql eval=FALSE}
dbExecute(con, "CREATE DATABASE IF NOT EXISTS dsaDB;")

```

### Initiate the database
Use the connection object to choose the database to manipulate.

```{r eval=FALSE}
dbExecute(con, "USE dsaDB;")

```

### Creating a table
The following example creates a table called *"Persons"* that contains five attributes: PersonID, LastName, FirstName, Address, and City:

The datatype parameter specifies the type of data the column can hold (e.g., varchar, integer, date, etc.)

```{sql eval=FALSE}
dbExecute(con, 
"CREATE TABLE IF NOT EXISTS Persons (
  PersonID int(11) NOT NULL AUTO_INCREMENT PRIMARY KEY,
  LastName varchar(255),
  FirstName varchar(255),
  Address varchar(255),
  City varchar(255)
);")

```


### Inserting records into a table
In order to add multiple rows into a table at once, use the following INSERT statement:

```{sql eval=FALSE}
dbExecute(con, 
"INSERT IGNORE INTO Persons (PersonID, LastName, FirstName, Address, City)
VALUES 
(1,'Erichsen','Tom ','1 Stavanger 4006','Oslo'),
(2,'Philips','James ','21 Folks Rd','Johannesburg'),
(3,'Lee','Kim ','2 Greg Rd','Nairobi'),
(4,'Doe','John ','55 Yawn Rd','Pretoria'),
(5,'Marks','Cindy ','22 Folks Rd','Abuja');")

```

The SELECT statement fetches all the records from the "Persons" table:

```{sql eval=FALSE}
dbGetQuery(con, "SELECT * FROM Persons;")

```


### Update records in a table

The UPDATE statement modifies the first customer (PersonID = 1) with a new last name and a new city.


```{sql eval=FALSE}
dbExecute(con, 
"UPDATE Persons
SET LastName = 'Schmidt', City= 'Frankfurt'
WHERE PersonID = 1;")

```


### Using aliases
The SELECT statement creates an alias for LastName column as *"Surname"*. Note: It requires double quotation marks if the alias name contains spaces:

```{sql eval=FALSE, connection=con}
dbGetQuery(con, "SELECT LastName AS 'Surname' FROM Persons;")

```




# Questions
1. Using an example, describe the difference in usage between the **RMariaDB** package library and the **DBI** package library.
2. What functionality has the **dbplyr** added to this tutorial?

# Exercise
This exercise is an assignment and you are required to:

a. Assume that a person can have one or more addresses, normalize the *"Persons"* in database **dsaDB** in order to capture this idea.

